Analyzing Network Transfer Protocols on a Simulated Interplanetary Internet

### 2024-04-10 paper/paper.tex
added data section

\usepackage{microtype}
\usepackage{biblatex}
\usepackage{hyperref}

@@

\begin{abstract}
  This paper discusses interplanetary networks, the challenges that can arise
  from them that make them a unique challenge compared to shorter-range computer
  networks, and the protocols that can be used in this type of network. In
  addition, we look at how one can use conventional network simulators to test
  interplanetary network implementations. We used the network simulator
  NS-3~\cite{ns-3} to test various transport layer protocols against each other
  in a network topology that we designed in the network simulator to emulate
  long-distance communications. Using this network simulation, we were able to
  compare UDP, TCP, and TCP New Reno protocols against each other to see how
  they each performed across such large distances with high delays.
\end{abstract}

@@

In their 2017 Masters Thesis, Md Monjurul Islam Khan~\cite{Khan2017} from
University of Manitoba wrote on how to create a satellite network simulation
based on ns3. They used a point to point ns3 module to create a network of 2
levels of orbiting satellites and fixed points on the earth which acts as
uplinks to the network. The paper analyzed TCP HighSpeed against TCP Vegas slow
start. However this paper stops at expanding the network beyond earth orbit, and
rather focuses on Earth orbit communication.

SNS3 is a simulation library which uses NS3 as a basis to create a framework
which to build simulations for geostationary satellites in the European
continent. It does not expand to an interplanetary network~\cite{Puttonen2014}.

@@

An interplanetary network is simply a network which where the nodes are
distributed across the surface or orbits of different planets. In the year of
2024 the current interplanetary is limited to Mars, Earth, the Moon, and
satellites orbiting the sun. The interplanetary network is also expanded to deep
space satellites, like the voyager probes. A sample message path across the
network could be that a message is passed from a on earth station through
satellite dish to a earth orbiting satellite, then when it has direct line of
sight with the Mars MAVEN sends the message. MAVEN then relays the message to a
Mars rover like Odyssey. This description hides the complexity which arises in
an interplanetary network.

In a traditional network a majority of the network will happen over a wired
connection. This means that the error is relatively low. This is no longer the
case in the case of satellites sending messages over potentially millions of
kilometers. Atmosphere in the Earth orbit, solar radiation, and dust cause
increased error in sending the message between entities. Inverse square law
dictates that over these large periods the energy needed to send the message so
that it can be read at its destination. Even the Doppler effect is needed to be
accounted for to tune into radio frequencies on the ISS. High error means that
sending a message can no longer be assumed to be sent reliably and has to be
compensated for. Beyond high error, large distances between entities and the
physical limitation that messages cannot travel faster than the speed of light.
Light, or any other energy takes 8 minutes to get from the Sun to Earth, and
thus for a round trip will take 16 minutes to complete a simple handshake, given
that the messages were sent correctly.

Other than physical limitations a interplanetary network has a rapidly changing
topology at anytime. For instance satellites which orbit planets can have their
field of view blocked as they move over the horizon. While terrestrial networks
also have rapidly changing networks at the edges, but the difference is that the
core of the network is stable traditionally. Where depending on the hour, day,
year, the core of the interplanetary network can change drastically. To be able
to connect to the network there are no well known places to access the network,
except for Earth.

To combat these challenges several protocols have been created. The SCPS TP is a
transport protocol which is used on the network layer. UDP is used in the lowest
reliable connections between nodes, and TCP is used in the most reliable network
conditions. When a reliable and efficient method of network transportation is
needed, then SCPS TP is used. It is a modified version of TCP which uses a slow
start protocol, reduced handshakes, reliable data transfer with reduced
confirmations. Other protocols like store and load, and Licklider are used to
also overcome these problems.

The rest of the paper remains to simulate these unique challenges, as well as
see in what conditions do the current protocols work.

@@

A significant challenge in creating a IPN is the rapidly changing topology. To
simulate these rapid changing conditions a physics engine was created to model
the orbits of several entities in the solar system. Several assumptions were
made in the simulator for the sake of simplicity as the goal of the simulator is
not to create a perfect model of the solar system and the entities inside of the
solar system, but rather conditions which a IPN would exist in.

The first assumption made was that all orbits are circular. Realistically,
orbits are oval shaped, with the two focal points. A result of this assumption
is the elimination of any difference in orbital speed due to Kepler's third law.
The orbital radius was chosen by taking the average orbital radius of the real
life celestial bodies. For example Earth's average orbital radius is 149.6
billion meters~\cite{nasa_earth}, and therefore orbits around the sun in a
circular orbit of radius 149.6 billion meters. Another result of this assumption
is that orbits are considered to be centered around the body which it orbits.
The Mars orbit has a focal point that is further in space, making it more oval
shaped than the earth orbit.~\cite{nasa_mars} While this could make changes to a
simulated network in some very specific situations, the primary behaviour of a
solar system is maintained, like retrograde orbits, periodical alignment,
predictable behaviour, and recursive orbits of sub orbiting object.

@@

Messages between entities were assumed to be sent at the speed of light in a
vacuum, as all forms of radio, lasers, or other wireless messages are forms of
electromagnetic radiation, and thus travel at the speed of light. Since the
propagation delay in an IPN is much higher than in terrestrial networks, other
delays such as queuing or processing delays were ignored as being negligible.

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. (See
appendix) All entities are able to block signals if the ray passes through the
radius of the entity. One of the largest factors contributing to transmission
error is the distance travelled by the electromagnetic signal between
transmitting and receiving nodes. By generalizing Friis' transmission
formula\cite{Friis} for isotropic emitters and receivers, we can obtain the
ratio of signal intensity received to signal intensity transmitted, known as
Free Space Path Loss (FSPL). We can use FSPL as our transmission error rate,
giving the following expression: $Error Rate = {(\frac{4 \pi d}{\lambda})}^2$
where $d$ is the distance between nodes in metres and $\lambda$ is the
wavelength of the electromagnetic wave used in transmission. As much of space
communications utilizes the Ka-band of frequencies (27--40
GHz)\cite{Morabito_Hastrup} we opted to use a transmission frequency of 30Ghz.
The transmission wavelength can be then calculated according to
$\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light. The
input of the physics engine is a time $t$ from some preset initial conditions
which outputs the state of all entities.

@@

\printbibliography{}

>>>>

\usepackage{microtype}
\usepackage{biblatex}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amsmath}

@@

\begin{abstract}
  This paper discusses interplanetary networks, the challenges that can arise
  from them that make them a unique challenge compared to shorter-range computer
  networks, and how our current network protocols can be applied in this application. In
  addition, we look at how one can use conventional network simulators to test
  interplanetary network implementations. We used the network simulator
  NS-3 ~\cite{ns-3} to test various transport layer protocols against each other
  in a network topology that we designed in the network simulator to emulate
  long-distance communications. Using this network simulation, we were able to
  compare UDP, TCP, and TCP New Reno protocols against each other to see how
  they each performed across such large distances with high delays. UDP preformed the 
  best with New Reno and TCP closely paired together.
\end{abstract}

@@

In their 2017 Masters Thesis, Md Monjurul Islam Khan~\cite{Khan2017} from
University of Manitoba wrote on how to create a satellite network simulation
based on NS-3. They used a point to point NS-3 module to create a network of 2
levels of orbiting satellites and fixed points on the earth which acts as
uplinks to the network. The paper analyzed TCP HighSpeed against TCP Vegas slow
start. However this paper stops at expanding the network beyond earth orbit, and
rather focuses on Earth orbit communication.

SNS3 is a simulation library which uses NS-3 as a basis to create a framework
which to build simulations for geostationary satellites in the European
continent. It does not expand to an interplanetary network~\cite{Puttonen2014}.

@@

An interplanetary network is simply a network which where the nodes are
distributed across the surface or in the orbits of different planets or other celestial objects.
In the year of
2024 the current interplanetary is limited to Mars, Earth, Earth orbit, Moon orbit, and
satellites orbiting the sun. The interplanetary network is also expanded to deep
space satellites, like the Voyager probes. A sample message path across the
network could be that a message is passed from a large network antenna on earth
to a earth orbiting satellite, then when it has direct line of
sight with MAVEN (a satellite orbiting mars) it sends the message. MAVEN then relays the message to a
Mars rover like Odyssey. This description hides the complexity which arises in
an interplanetary network which may seem like a extremely simple network.

In a traditional network a majority of the network traffic will occur over a wired
connection, especially for transfers of large amounts of data.
This means that the error is relatively low. This is no longer the
case in the case of satellites sending messages over potentially millions of
kilometers. Atmosphere in the Earth orbit, solar radiation, and dust cause
increased error in sending the message between entities. Inverse square law
dictates that over these large periods the energy needed to send the message so
that it can be read at its destination. Even the Doppler effect is needed to be
accounted for to tune into radio frequencies on the ISS. ~\cite{qslCompensatingDoppler} High error means that
sending a message can no longer be assumed to be sent reliably and has to be
compensated for. Beyond high error, large distances between entities and the
physical limitation that messages cannot travel faster than the speed of light.
Light, or any other energy takes approximately 8 minutes to get from the Sun to Earth, ~\cite{nasa_earth} and
thus for a round trip will take 16 minutes to complete a simple handshake, given
that the messages were sent correctly.

Other than physical limitations a interplanetary network has a rapidly changing
topology at anytime. For instance satellites which orbit planets can have their
field of view blocked as they move over the horizon. While terrestrial networks
also have rapidly changing networks at the edges, but the difference is that the
core of the network is often stable. Where depending on the hour, day,
year, the core of the interplanetary network can change drastically. To be able
to connect to the network there are no well known places to access the network,
except for Earth.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{media/ipn.png}
  \caption{A sample of the current interplanetary network}
\end{figure}

To combat these challenges several protocols have been created. The SCPS TP is a
transport protocol which is used on the network layer. UDP is used in the lowest
reliable connections between nodes, and TCP is used in the most reliable network
conditions. When a reliable and efficient method of network transportation is
needed, then SCPS TP is used. It is a modified version of TCP which uses a slow
start protocol, reduced handshakes, reliable data transfer with reduced
confirmations. Other protocols like store and load, and Licklider are used to
also overcome these problems. In this paper since no complete implementation of 
SCPS TP has been created for NS-3, New Reno TCP was used instead.

The rest of the paper remains to simulate these unique challenges, as well as
see how protocols operate at these different situations.

@@

A significant challenge in modeling a IPN is modeling the rapidly changing topology. To
simulate these rapid changing conditions a physics engine was created to model
the orbits of several entities in the solar system. Several assumptions were
made in the simulator for the sake of simplicity as the goal of the simulator is
not to create a perfect model of the solar system and the entities inside of the
solar system, but rather conditions which a IPN would exist in.

The first assumption made was that all orbits are circular. Realistically,
orbits are oval shaped, with two focal points. A result of this assumption
is the elimination of any difference in orbital speed due to Kepler's third law.
Normally a planet will accelerate or decelerate depending on where it is in 
its orbit around the sun. 
The orbital radius was chosen by taking the average orbital radius of the real
life celestial bodies. For example Earth's average orbital radius is 149.6
billion meters~\cite{nasa_earth}, and therefore orbits around the sun in a
circular orbit of radius 149.6 billion meters in our model. Another result of this assumption
is that orbits are considered to be centered around the body which it orbits.
The Mars orbit has a focal point that is further in space, making it more oval
shaped than the earth orbit.~\cite{nasa_mars} While this could make changes to a
simulated network in some very specific situations, the primary behaviour of a
solar system is maintained, like retrograde orbits, periodical alignment,
predictable behaviour, and recursive orbits of sub orbiting object.

@@

Messages between entities were assumed to be sent at the speed of light in a
vacuum, as all forms of radio, lasers, or other wireless messages are forms of
electromagnetic radiation, and thus travel at the speed of light. Since the
propagation delay in an IPN is much higher than in terrestrial networks, other
delays such as processing delays were ignored as being negligible.

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. (See
appendix) All entities are able to block signals if the ray passes through the
radius of the entity. One of the largest factors contributing to transmission
error is the distance travelled by the electromagnetic signal between
transmitting and receiving nodes. By generalizing Friis' transmission
formula\cite{Friis} for isotropic emitters and receivers, we can obtain the
ratio of signal intensity received to signal intensity transmitted, known as
Free Space Path Loss (FSPL). We can use FSPL as our transmission error rate,
giving the following expression: $Error Rate = {(\frac{4 \pi d}{\lambda})}^2$
where $d$ is the distance between nodes in metres and $\lambda$ is the
wavelength of the electromagnetic wave used in transmission. As much of space
communications utilizes the Ka-band of frequencies (27--40
GHz)\cite{Morabito_Hastrup} we opted to use a transmission frequency of 30Ghz.
The transmission wavelength can be then calculated according to
$\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/sun_orbit.png}
  \includegraphics[width=0.4\textwidth]{media/moon_orbit.png}
  \caption{A visual output of the physics model. Left: Mars and Earth orbiting the sun. Right: The ISS (overlapping the center) and the Moon orbiting the Earth}
\end{figure}

The physics simulation modeled the Sun, Earth, Moon, International Space Station,
Mars, and the Mars Orbiter. All were able to connect to the network except for the 
the Sun and the Moon. The input of the physics engine is a time $t$ from some preset initial conditions
which outputs the state of all entities. Which is fed into the NS-3 simulation.

@@

During the work on this project we contributed to the open source NS-3 project 
by committing a change to fix a bug which occurred when running the program on 
different operating systems.

\section{Data}
The NS-3 simulation ran on the basis of transferring data between Earth and 
Mars.

Data was scraped from the trace output that was generated while running the 
ns3 simulation. Trace data was given when a message was queued, dequeued, 
and received at the destination. This data also included the amount of 
data being sent in bytes. There is other data on the messages being sent, but 
was not tracked for the purpose of this project. The trace data was parsed 
using regex expressions to extract the needed numbers from the string.
This data was then fed into a pandas dataframe. A unique id had to be 
generated for each message since once a message was recieved the id was 
reused. This was done by tracking when each message was received with id 
$x$ and after incrementing a number by one, say $y$. The unique id was then 
created by hashing $xy$. Once the data frame was created it had to be merged 
on itself to have one row of the dataframe represent exactly one message rather 
then one row be either queue, dequeue, or received.

Time from being queued to received, time being queued, and message time in transit
were all calculated based off of the data.

\[\text{total\_package\_time} = \text{time\_recieved} - \text{time\_queued}\]

\[\text{time\_queued} = \text{time\_dequeued} - \text{time\_queued}\]

\[\text{total\_time} = \text{time\_received} - \text{time\_dequeued}\]

After the data was processed it was saved in a csv file with one row representing 
exactly one message that was sent over the network. This file was close to 
2.5 GB in size.

Statistics were calculated on the entire column of the calculated rows, and 
bytes sent. Messages were also grouped together which occurred during a similar 
time window and further averages were calculated to find how the average changed 
over time of the simulation. These calculations are shown in the graphs in the 
analysis section.

\section{Analysis}

\subsection{UDP}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/udp.png}
  \includegraphics[width=0.4\textwidth]{media/udp_data.png}
  \caption{UDP analyzed}
\end{figure}

UDP had an average time from being queued to received of
195.4 seconds, an average time being queued of
18.36 seconds, and an average message time in transit
177.0 seconds. The average amount of data sent was 
1051.6 bytes. Notice the scientific notation in the 
corner of the graph. The graph shows a large amount of 
sensitivity to the motion of the planets which can be 
seen by the curvature of the graph. At time around 28000
seconds, the topology suddenly changed. Caused by a satellite
blocking the path between Mars and Earth. This caused a secondary, 
less efficient path to be taken. This shows why there are several 
"layers" of time which it took. They likely represent the different paths 
which the message took before being sent to its destination. Each quanta in
the time it took to complete the message 
represents the smaller jump it takes to get from a satellite to its 
orbiting entity. The increase in time likely represents that while the 
direct Earth and Mars link was down the smaller satellites were actually orbiting 
away from the message destination.

\subsection{TCP}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/tcp.png}
  \includegraphics[width=0.4\textwidth]{media/tcp_data.png}
  \caption{UDP analyzed}
\end{figure}

\printbibliography
